### 2018-01-23 submission/manuscript.Rmd
update abstract

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness resulted in small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only resulted in an increased relative risk for carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern, with their presence significantly increasing the relative risk for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value = `r format(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]), digits = 2)`). Using the whole community versus only CRC-associated genera to build a prediction model resulted in higher classification success, based on Area Under the Curve (AUC), for both adenoma and carcinoma using fecal and tissue samples. The most important OTUs for these models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Overall, there were less associations between the microbiota and adenoma and one reason why this may be is that most studies were only adequately powered for large effect sizes.

>>>>

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness resulted in small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only resulted in an increased relative risk for carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Mouth-associated genera were consistently in the top 5 OTUs or genera with the highest relative risk of adenoma and carcinoma for both stool and tissue. Using the whole community versus selected generea with the highest and lowest relative risk, to build a prediction model, resulted in similar classification success, based on Area Under the Curve (AUC), for both adenoma and carcinoma using fecal and tissue samples. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Overall, there were less associations between the microbiota and adenoma and one reason why this may be is that most studies were only adequately powered for large effect sizes.

### 2018-01-23 submission/manuscript.tex
update abstract

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness resulted in small
increases in relative risk for adenoma (P-value = 0.032) and carcinoma
(P-value = 0.00034) while the reduction in Shannon diversity only
resulted in an increased relative risk for carcinoma (P-value = 0.0047).
Previously associated colorectal cancer genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
followed a similar pattern, with their presence significantly increasing
the relative risk for carcinoma (P-value \textless{} 0.05) but not
adenoma (P-value check tables 0.05) with the exception of
\emph{Porphyromonas} (P-value = 0.023). Using the whole community versus
only CRC-associated genera to build a prediction model resulted in
higher classification success, based on Area Under the Curve (AUC), for
both adenoma and carcinoma using fecal and tissue samples. The most
important OTUs for these models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Overall, there were less associations between the microbiota
and adenoma and one reason why this may be is that most studies were
only adequately powered for large effect sizes.

>>>>

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness resulted in small
increases in relative risk for adenoma (P-value = 0.032) and carcinoma
(P-value = 0.00034) while the reduction in Shannon diversity only
resulted in an increased relative risk for carcinoma (P-value = 0.0047).
Mouth-associated genera were consistently in the top 5 OTUs or genera
with the highest relative risk of adenoma and carcinoma for both stool
and tissue. Using the whole community versus selected generea with the
highest and lowest relative risk, to build a prediction model, resulted
in similar classification success, based on Area Under the Curve (AUC),
for both adenoma and carcinoma using fecal and tissue samples. The most
important groups within the full community models consistently belonged
to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
\emph{Roseburia} across studies. Overall, there were less associations
between the microbiota and adenoma and one reason why this may be is
that most studies were only adequately powered for large effect sizes.
